waRne - Putting cricket analytics in a spin

### 2020-10-01 James_paper_copy.Rmd
adding my copy of the paper with the graphs

---
title: "waRne - Putting cricket analytics in a spin"
authors: Robert(1), James (2)
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
bibliography: references.bib

---

@@

@@

The importance of reproducibility, and the related issue of open access to data, has received a lot of recent attention. Momentum on these issues is gathering in the sports analytics community. While cricket is the second largest commercial sport in the world, unlike other popular international sports, there has been no mechanism for the public to access comprehensive statistics on players and teams. Expert commentary currently relies heavily on data that isn't made readily accessible and this produces an unnecessary barrier for the development of an inclusive sports analytics community.

@@

Cricket is the second largest sport in the world. However, unfortunetaly there is no easy accessible way to access ball by ball data nor aggregated statistics of teams and players. Data while available on sites like [espncricinfo](https://www.espncricinfo.com/) are not in an easy to use form. For example, each match is listed on different webpages so hours upon hours of time would be required to copy and paste a single season. Hence, there are significant logistical barriers for prospective fans and analysts studying the game, which stagnats understanding of cricket.

This paper describes the waRne package, the first to provide free and easy access to data for cricket for fans. Web scraping tools are avaiable for fans to easily scrape the play by play commentary data on espncricinfo. For the first time fans can evaluate their favourite teams and players and do so in a reproducible and accessible manner.

@@

Data accessibility enables fans, analyts and researchers to better understand the game. Through being able to reproduce common popular metrics, visualisations and article findings.

@@

For example in this [series of tweets](https://twitter.com/benjonescricket/status/1219898578719723520) we see the narrative being pushed that Steve Smith is a good player vs pace bowling unfortunetly just a few months prior the same company and journalist published an article which had Steve Smith doing much worse against pace (balls above 140km/h). Unlike a similar sport baseball, fans have no easily accisible way of seeing if Steve Smith vs pace is a strength as alluded to in the original tweet, or a weakness like the same persons published online article. In comparsion, fans are able to get a breakdown of Mike Trout vs fastballs from using [baseballr](https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/) which provides access through statcast data from [baseballsavant]([https://baseballsavant.mlb.com)](https://baseballsavant.mlb.com)) .

@@

However though using waRne we are able to not only compare Steve Smiths performance vs bowling type but we are able to easily compare Steve Smiths performance vs bowling types with other players.

@@

```{r}
library(tidyverse)

# steve smith vs pace graph

@@

# insert above plots here
```

@@

>>>>

---
title: "waRne - Putting cricket analytics in a spin"
bibliography: references.bib
output:
  pdf_document:
    toc: yes
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
authors: Robert(1), James (2)
---

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(stringr)
library(curl)
library(tidyverse)
library(magrittr)
library(lubridate)
library(ggplot2)
library(ggalt)
library(gganimate)
library(gifski)
player_database <- 
  read_csv("/Users/James/James/Cricket_Thesis/Data/new_player_database_crickjam.csv") %>% 
  dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
  distinct() %>% 
  dplyr::filter(player_name != "Rohit Gurunath Sharma") %>% 
  bind_rows(read_csv("/Users/James/James/Cricket_Thesis/Data/new_player_database_crickjam.csv") %>% 
              dplyr::filter(player_id == 34102) %>% 
              dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
              drop_na() %>% 
              distinct()) %>% 
  dplyr::filter(player_name !="Sultan Ahmed")

bowler_vars <- 
  player_database %>% 
  dplyr::select(-c(batting_style, player_name)) %>% 
  drop_na()%>% 
  mutate(bowling_style = recode(bowling_style, "Right-arm fast" = "Pace",
                                "Slow left-arm orthodox"= "Spin",
                                "Legbreak"= "Spin",
                                "Right-arm medium"= "Pace",
                                "Left-arm medium-fast"= "Pace",
                                "Right-arm offbreak"= "Spin",
                                "Legbreak googly"= "Spin",
                                "Left-arm fast"= "Pace",
                                "Right-arm medium-fast"= "Pace",
                                "Left-arm fast-medium"= "Pace",
                                "Right-arm fast-medium"= "Pace",
                                "Left-arm medium"= "Pace",
                                "Slow left-arm chinaman"= "Spin",
                                "Right-arm offbreak, Legbreak"= "Spin",
                                "Right-arm medium, Right-arm offbreak"= "Spin",
                                "Left-arm medium, Slow left-arm orthodox, 
                                               Slow left-arm chinaman"= "Spin", 
                                "Left-arm medium, Slow left-arm orthodox, Slow left-arm chinaman" = "Spin"))

batting_vars <- 
  player_database %>% 
  dplyr::select(-c(bowling_style, player_name)) %>%
  drop_na()

cricjam_data <- 
  read_csv("/Users/James/James/Reproducibility-in-Cricket/Data/combined_2020_09_11.csv") %>% 
  dplyr::filter(!grepl("No result", match_summary)) %>%
  mutate(bucketed_year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
  left_join(batting_vars, by = c("batsman_id" = "player_id")) %>% 
  left_join(bowler_vars, by = c("bowler_id" = "player_id")) %>% 
  mutate(batsman_name = str_remove(batsman_name, "\\(\\)"), 
         batsman_name = str_trim(batsman_name), 
         runs = extra_runs+earned_runs)
# 
# cricjam_data_partner <- 
#   cricjam_data %>% 
#   dplyr::select(match_id, inning, batsman_name, batsman_id, batting_order) %>% 
#   distinct() %>% 
#   arrange(batting_order) %>% View()

wicket_types <-
  cricjam_data %>% 
  dplyr::filter(wicket==TRUE) %>% 
  dplyr::select(wicket_description, play_description, bowler, batsman, bowler_name, batsman_name, 
                match_id, inning, over) %>%
  mutate(caught_out = if_else(str_detect(wicket_description, " c "), TRUE, FALSE), 
         bowled_out = if_else(str_detect(wicket_description, " b "), TRUE, FALSE), 
         run_out = if_else(str_detect(wicket_description, " run out "), TRUE, FALSE), 
         stumped_out = if_else(str_detect(wicket_description, " st "), TRUE, FALSE), 
         leg_before_wicket = if_else(str_detect(wicket_description, " lbw "), TRUE, FALSE), 
         retired_hurt = if_else(str_detect(wicket_description, " retired hurt "), TRUE, FALSE), 
         obstructing_the_field = if_else(str_detect(wicket_description, " obstructing the field "), TRUE, FALSE), 
         retired_out = if_else(str_detect(wicket_description, " retired out "), TRUE, FALSE), 
         null = if_else(str_detect(wicket_description, " null "), TRUE, FALSE)) %>% 
  # all_conditions = if_else(caught_out == TRUE |
  #                          bowled_out == TRUE | caught_out == TRUE |
  #                            run_out == TRUE |retired_out == TRUE |
  #                            retired_hurt == TRUE |leg_before_wicket == TRUE |
  #                            obstructing_the_field == TRUE |null == TRUE |
  #                            stumped_out == TRUE, TRUE, FALSE))
  mutate(out_condition = case_when(caught_out==TRUE |
                                     caught_out==TRUE & bowled_out == TRUE ~ "Caught", 
                                   run_out == TRUE |
                                     run_out==TRUE & bowled_out == TRUE ~ "Run Out", 
                                   stumped_out == TRUE |
                                     stumped_out==TRUE & bowled_out == TRUE ~ "Stumped", 
                                   leg_before_wicket == TRUE |
                                     leg_before_wicket==TRUE & bowled_out == TRUE ~ "Leg Before Wicket", 
                                   retired_hurt == TRUE |
                                     retired_hurt==TRUE & bowled_out == TRUE ~ "Retired Hurt", 
                                   obstructing_the_field == TRUE |
                                     obstructing_the_field==TRUE & bowled_out == TRUE ~ "Obstructing the Field", 
                                   retired_out == TRUE |
                                     retired_out==TRUE & bowled_out == TRUE ~ "Retired Out", 
                                   null == TRUE |
                                     null==TRUE & bowled_out == TRUE ~ "Caught", 
                                   bowled_out == TRUE & caught_out == FALSE |
                                     bowled_out == TRUE & stumped_out == FALSE |
                                     bowled_out == TRUE & run_out == FALSE |
                                     bowled_out == TRUE & leg_before_wicket == FALSE |
                                     bowled_out == TRUE & retired_hurt == FALSE |
                                     bowled_out == TRUE & obstructing_the_field == FALSE |
                                     bowled_out == TRUE & retired_out == FALSE |
                                     bowled_out == TRUE & null == FALSE ~"Bowled"), 
         out_batsman = case_when(out_condition == "Bowled" ~ word(wicket_description, 1, sep = " b "), 
                                 out_condition == "Caught" & null == FALSE ~ word(wicket_description, 1, sep = " c "), 
                                 out_condition == "Caught" & null == TRUE ~ word(wicket_description, 1, sep = " null "), 
                                 out_condition == "Run Out" ~ word(wicket_description, 1, sep = " run out "), 
                                 out_condition == "Stumped" ~ word(wicket_description, 1, sep = " st "), 
                                 out_condition == "Leg Before Wicket" ~ word(wicket_description, 1, sep = " lbw "), 
                                 out_condition == "Retired Hurt" ~ word(wicket_description, 1, sep = " retired hurt "), 
                                 out_condition == "Obstructing the Field" ~ word(wicket_description, 1, sep = " obstructing the field "), 
                                 out_condition == "Retired Out" ~ word(wicket_description, 1, sep = " retired out ")),  
         out_batsman = str_trim(out_batsman)) %>% 
  dplyr::select(-c(caught_out, bowled_out, stumped_out, leg_before_wicket, retired_hurt, 
                   obstructing_the_field, retired_hurt, null, run_out, retired_out)) %>% 
  mutate(batsman_or_partner_out = if_else(out_batsman == batsman_name, "Batsman","Partner"))

cricjam_data <- 
  cricjam_data %>% 
  left_join(wicket_types, by = c("wicket_description" = "wicket_description", 
                                 "play_description" = "play_description", 
                                 "bowler" = "bowler", 
                                 "batsman" = "batsman", 
                                 "batsman_name" = "batsman_name", 
                                 "match_id" = "match_id", 
                                 "inning" = "inning", 
                                 "over" = "over", 
                                 "bowler_name" = "bowler_name")) %>% 
  rename(inning_summary = inning_total)

only_big_t20_matches <- 
  cricjam_data %>% 
  dplyr::filter(tournament_type == "Twenty20", 
                tournament_name %in% c("Twenty20 Big Bash", "Indian Premier League", "Big Bash League", 
                                       "Champions League Twenty20", "Pepsi Indian Premier League")) %>% 
  dplyr::select(match_id) %>% 
  distinct() %>% 
  pull(match_id)

cricjam_data_big <- 
  cricjam_data %>% 
  dplyr::filter(tournament_type == "Twenty20 Internationals") %>% 
  bind_rows(cricjam_data %>% 
              dplyr::filter(match_id %in% only_big_t20_matches))
```

\color{red}{\text{Yeah I forgot I wouldn't have any internet on the plane, so everything I wrote is not referencing anything from a link or package I didn't have downloaded prior to boarding}}

@@

The importance of reproducibility, and the related issue of open access to data, has received a lot of recent attention. Momentum on these issues is gathering in the sports analytics community, particularly in North America where open access to quality sports data has become the norm. Cricket is the second largest commercial sport in the world, but unlike other popular international sports, there has been no mechanism for the public to access comprehensive statistics on players and teams. Expert commentary currently relies heavily on data that isn't made readily accessible and this produces an unnecessary barrier for the development of an inclusive sports analytics community.

@@

Cricket is the second largest sport in the world. However, unfortunetaly there is no easy accessible way to access ball by ball data nor aggregated statistics of teams and players. Data while available on sites like [espncricinfo](https://www.espncricinfo.com/) are not in an easy to use form. For example, each match is listed on different webpages so hours upon hours of time would be required to copy and paste a single season. Hence, there are significant logistical barriers for prospective fans and analysts studying the game, which stagnats understanding of cricket. The Duckworth-Lewis-Stern table, used to reset targets for matches that experience rain delays in all of cricket, is maintained by one man in Australia. Not even teams have access to the table. \color{red}{text{what do they do, call it into the ICC to get a temporary copy or something when there's a delay?}}

This paper describes the waRne package, the first to provide free and easy access to data for cricket for fans. Web scraping tools are avaiable for fans to easily scrape the play by play commentary data on espncricinfo. For the first time fans can evaluate their favourite teams and players and do so in a reproducible and accessible manner. We have included the data for ODI and T20 from the 2010 archived season on ESPCcricinfo up to the matches on September 8, 2020. If a person wishes to expand this data to include more recent games, this package can scrape and update the dataset.

@@

Data accessibility enables fans, analysts and researchers to better understand the game. Through being able to reproduce common popular metrics, visualisations and article findings.

@@

\color{red}{\text{I'm not sure if we should slam the datajam competition directly. I think the guy tweeting out conflicting news is a good point and I want to keep that in because he was a dick about his results, but we just lost to Quantium and I worry it will make us seem petty or look like sore losers. We could just indirectly show that they were wrong, but not mention them at all, I would feel fine with that}}

For example in this [series of tweets](https://twitter.com/benjonescricket/status/1219898578719723520) we see the narrative being pushed that Steve Smith is a good player vs pace bowling unfortunetly just a few months prior the same company and journalist published an article which had Steve Smith doing much worse against pace (balls above 140km/h). Unlike a similar sport baseball, fans have no easily accisible way of seeing if Steve Smith vs pace is a strength as alluded to in the original tweet, or a weakness like the same persons published online article. In comparsion, fans are able to get a breakdown of Mike Trout vs fastballs from using [baseballr](https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/) which provides access through statcast data from [baseballsavant]([https://baseballsavant.mlb.com)](https://baseballsavant.mlb.com)).

@@

With the commentary data on cricinfo, we are not able to get the same resolution on delivery speed and location as the statcast data provides in baseball, but we can often learn the type of delivery the batsman faces through the years. This represents an enormous leap in publically available data, and we hope will lead to the eventual release of the ball tracking data used butnot shared in crickets leagues around the world

Armed with this new dataset and the waRne package, we are able to not only compare Steve Smiths performance vs bowling type but we are able to easily compare Steve Smiths performance vs bowling types with other players, across multiple seasons, leagues, and cricket formats.

@@

\color{red}{\text{So the way the scraper is set up, I grab the play links when scraping the scorecard and commentary along with the id and name, and then run a separate scraper for the batsman/bowler profiles. So they are separate datasets that have to be joined together, and we drop mostly everything except batting and bowling style. I can probably make an all in one scraper, but it'll be a bit tricky and if we want to put this out Friday it might be best to leave them separate and just walk people through it in the readme. Otherwise it might not work at all, and the scraper already breaks down so often you have to pay attention more than you should}}

cricjam_data_big %>% 
	dplyr::filter(batsman_id == 267192, 
								!is.na(bowling_style)) %>% 
	group_by(bowling_style, earned_runs) %>% 
	summarise(total = n()) %>% 
	group_by(bowling_style) %>% 
	mutate(totals = sum(total),
				 percent = total/totals, 
				 earned_runs = as.factor(earned_runs)) %>% 
	ggplot(aes(y = bowling_style, x = percent, color = earned_runs))+
	geom_point() +
	labs(color = "Runs", 
			 x = "% Balls Faced", 
			 y = "Bowling Style", 
			 title = "Steve Smith's Distribution of Runs Against Spin and Pace")

@@

cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result, year) %>% 
	summarise(win_count = n()) %>% 
	ungroup() %>% 
	pivot_wider(names_from = batting_team_result, values_from = win_count) %>% 
	rename(superover = 'super-over') %>% 
	replace_na(list(superover=0)) %>% 
	mutate(total_games = lose+win+superover, 
				 win_percentage = win/total_games, 
				 year = as.factor(year)) %>% 
	ggplot(aes(x = year, y = win_percentage, group = 1)) +
	geom_line(color = 'red')

@@

dat <- 
	cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result) %>% 
	summarise(win_count = n()) %>% 
	dplyr::filter(batting_team_result != 'super-over')

N = sum(dat$win_count)
wins = dat$win_count[2]

z = (wins/N-0.5)/sqrt((0.5^2)/N)

1-pnorm(abs(z))

# No eveidence to suggest the 2 inning team wins more
